---
title: "Make a web app with Shiny"
layout: lab
tags:
- R
hidden: true
output: html_document
---

Today we will learn how to make and deploy interactive web applications with [Shiny](http://shiny.rstudio.com/).  Shiny allows you to display your R analysis on the web to anyone.  For example, here is an application that I wrote to visualize a [Markov Chain simulation of genetic drift](http://symposium.plb.ucdavis.edu:3838/MarkovDrift/).  A student from last years BIS180L is now working in my lab and is working up a [visualizer for QTL data and gene expression](http://symposium.plb.ucdavis.edu:3838/QTL-Visualization/).  The Shiny website has plenty of [additional examples](http://shiny.rstudio.com/gallery/).

The [tutorial at Rstudio](http://shiny.rstudio.com/tutorial/) is good but pretty involved.  I have tried to distill some of the prinicples below; if you like it and want more then feel free to do the [tutorial](http://shiny.rstudio.com/tutorial/) as well.

## Teams

Today we will work in teams of three.  Each team will produce and deploy a Shiny app that will be collectively graded.

| Student 1              | Student 2              | Student 3      | Repository                 |
|------------------------|------------------------|----------------|----------------------------|
| Tyler Kent             | Gurkern Sufi           | Chadwick Ho    | Shiny_Tyler.Sufi.Chad      |
| Samuel Deck            | Ed Sung                | Nabeel Sami    | Shiny_Sam.Ed.Nabeel        |
| Derek Conkle-Gutierrez | Maria I Chacon Sanchez | Kevin Hu       | Shiny_Derek.Maria.Kevin    |
| Anna Alekina           | Kristina Bow           | Robert Rider   | Shiny_Anna.Kristina.Bobby  |
| Lewis Roa              | Jenny Cade             | Daniel Ahrens  | Shiny_Lewis.Jenny.Daniel   |
| Matthew Denny          | Derek Lin              | Marina Becker  | Shiny_Matthew.Derek.Marina |
| Eric Wetzel            | Evan Adams             | Calvin Mackey  | Shiny_Eric.Evan.Calvin     |
| Scot Nicol             | Patrick Mahle          | Anjali Kadakia | Shiny_Scot.Patrick.Anjali  |
| Heesun Cha             | Khoa Nguyen            | Andy Peng      | Shiny_Sunny.Khoa.Andy      |
| Leslie Herrera         | Pallavi Setty          |                | Shiny_Leslie.Pallavi       |


## Preliminaries

Make sure that you have the latest version of Shiny and Shinyapps

```{r prelim, eval=FALSE}
install.packages(shiny)
devtools::install_github('rstudio/shinyapps') #this is only needed for deployment on the shinyapp web server.
```


## Components of a Shiny App

A ShinyApp consists of two R scripts:

* __ui.R__ This script controls the user interface (i.e. the design of the webpage, the input and the output).
* __server.R__ This script does the work of performing any analysis, creating graphs, and creating tables

These two scripts must be saved together in a single directory.  Each app much be saved in a different directory.

## ui.R

Here is a sample ui.R script.  This script makes use of R's build in data set on iris.  For more information on this data set you can type `?iris` in R.

``` {r, eval=FALSE}
library(shiny)

# Define UI for application that draws a histogram
shinyUI(fluidPage( #create the overall page

  # Application title
  titlePanel("Iris Data"),

  # Sidebar with a radio box to input which trait will be plotted
  sidebarLayout(
    sidebarPanel(
      radioButtons("trait", #the input variable that the value will go into
                  "Choose a trait to display:",
                  c("Sepal.Length",
                    "Sepal.Width",
                    "Petal.Length",
                    "Petal.Width")
    ),

    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("boxPlot")
    )
  )
))
```

To see all the types of input and output that can be included, see the [Shiny reference](http://shiny.rstudio.com/reference/shiny/latest/)





## Scoring (out of 20 points)

14 points for (reasonable) code, even if it doesn't run.  
+ 2 points for functional web app.  
+ 2 points for using two or more input types.  
+ 2 points for good annotation on the web page (a new user would understand what the app is about).  

## Going further

You can also include Shiny apps [directly in your Rmarkdown files](http://rmarkdown.rstudio.com/authoring_shiny.html) to be rendered into html.  In some ways the syntax is even easier because you only need a single file.  You can also include interactive graphs in slide presentations this way.